Can We Formalise Type Theory Intrinsically without Any Compromise? A Case Study in Cubical Agda

### 2025-12-04 tex/CPP26/final/main.tex
ACM cannot deal with subtitles together with question marks

\title{Can We Formalise Type Theory Intrinsically without Any Compromise?}
\subtitle{A Case Study in Cubical Agda}

@@

Finally, we introduce the code \ensuremath{\Varid{π}} for \ensuremath{\Conid{Π}} and the type equality \ensuremath{\Conid{Elπ}} to complete our definition of type theory using QIIRT:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π}\;{}\<[8]%
\>[8]{}\mathbin{:}\;{}\<[11]%
\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π[]}\;{}\<[8]%
\>[8]{}\mathbin{:}\;{}\<[11]%
\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;{}\<[35]%
\>[35]{}\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;{}\<[35]%
\>[35]{}\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}\highlight{addition}{(\Varid{pa'}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{a}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}\highlight{addition}{(\Varid{pb'}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑El}\;\kern-1.5pt]_{\text{t}})\;{}\<[43]%
\>[43]{}\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}(\Varid{π}\;\Varid{a}\;\Varid{pa}\;\Varid{b}\;\Varid{pb})\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;{}\<[E]%
\\
\>[8]{}\mathop\equiv\;{}\<[11]%
\>[11]{}\Varid{π}\;(\Varid{a}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{pa'}}\;(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑El}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{pb'}}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{tyOf}\;(\Varid{π}\;\anonymous \;\anonymous \;\anonymous \;\anonymous )\;\mathrel{=}\;\Conid{U}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{Elπ}\;\mathbin{:}\;(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{→}\;(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{→}\;\Conid{El}\;(\Varid{π}\;\Varid{a}\;\Varid{pa}\;\Varid{b}\;\Varid{pb})\;\highlight{addition}{\Varid{refl}}\;\mathop\equiv\;\Conid{Π}\;(\Conid{El}\;\Varid{a}\;\Varid{pa})\;(\Conid{El}\;\Varid{b}\;\Varid{pb}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{tyOf}\;(\Varid{π[]}\;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\Varid{i})\;\mathrel{=}\;\Conid{U[]}\;\Varid{i}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

For example, the coherence proof for \ensuremath{\Varid{ηπ}} is given by
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{beginSub[}\;\Varid{ηπ}\;\mskip1.5mu]\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{elimSub}\;\Varid{σ}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{≡Sub[}\;\Varid{ηπ}\;\Varid{]⟨}\;\Varid{ηπ∙}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{⟩}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π₁}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{,}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π₂}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{tyOfπ₂}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\kern-1.5pt]^{\bullet}_{\text{t}}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{≡Sub[}\;\Varid{refl}\;\Varid{]⟨}\;\Varid{cong}\;(\Varid{λ}\;\Varid{z}\;\Varid{→}\;\Varid{...}\;\Varid{,}\;\Varid{...}\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{z}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;(\Conid{Ty∙-is-set}\;{}\<[71]%
\>[71]{}\anonymous \;\anonymous )\;\Varid{⟩}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π₁}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{,}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{elimTm}\;(\Varid{π₂}\;\Varid{σ})\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{elimTyOf}\;(\Varid{π₂}\;\Varid{σ})\;\Varid{refl}\;\kern-1.5pt]^{\bullet}_{\text{t}}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{∎}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
with the steps \ensuremath{\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{p}\;\Varid{]⟨}\;\Varid{p}^{\bullet}\;\Varid{⟩}\;\Varid{τ}^{\bullet}} implemented using \ensuremath{\_\Varid{∙}_{\mathsf{P}}\_}.
These steps give us an equation over \ensuremath{\Varid{ηπ}\;\Varid{∙}\;(\Varid{refl}\;\Varid{∙}\;\Varid{refl})}, and \ensuremath{\Varid{beginSub[}\;\Varid{ηπ}\;\Varid{]\char95 }} gives us an easy way to correct this to an equation over \ensuremath{\Varid{ηπ}} as desired instead.

>>>>

\title[Can We Formalise Type Theory Intrinsically without Any Compromise?]{Can We Formalise Type Theory Intrinsically without Any Compromise? A Case Study in Cubical Agda}

@@

Finally, we introduce the code \ensuremath{\Varid{π}} for \ensuremath{\Conid{Π}} and the type equality \ensuremath{\Conid{Elπ}} to complete our definition of type theory using QIIRT:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π}\;{}\<[8]%
\>[8]{}\mathbin{:}\;{}\<[11]%
\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π[]}\;{}\<[8]%
\>[8]{}\mathbin{:}\;{}\<[11]%
\>[11]{}(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;{}\<[35]%
\>[35]{}\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}\highlight{addition}{(\Varid{pa'}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{a}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}\highlight{addition}{(\Varid{pb'}\;\mathbin{:}\;\Varid{tyOf}\;(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑El}\;\kern-1.5pt]_{\text{t}})\;{}\<[43]%
\>[43]{}\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[8]{}\Varid{→}\;{}\<[11]%
\>[11]{}(\Varid{π}\;\Varid{a}\;\Varid{pa}\;\Varid{b}\;\Varid{pb})\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}}\;{}\<[E]%
\\
\>[8]{}\mathop\equiv\;{}\<[11]%
\>[11]{}\Varid{π}\;(\Varid{a}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{pa'}}\;(\Varid{b}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{↑El}\;\kern-1.5pt]_{\text{t}})\;\highlight{addition}{\Varid{pb'}}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{tyOf}\;(\Varid{π}\;\anonymous \;\anonymous \;\anonymous \;\anonymous )\;\mathrel{=}\;\Conid{U}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\anonymous \;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{Elπ}\;\mathbin{:}\;(\Varid{a}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\highlight{addition}{(\Varid{pa}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{a}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{→}\;(\Varid{b}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{El}\;\Varid{a}\;\Varid{pa}))\;\highlight{addition}{(\Varid{pb}\;\mathbin{:}\;\Varid{tyOf}\;\Varid{b}\;\mathop\equiv\;\Conid{U})}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{→}\;\Conid{El}\;(\Varid{π}\;\Varid{a}\;\Varid{pa}\;\Varid{b}\;\Varid{pb})\;\highlight{addition}{\Varid{refl}}\;\mathop\equiv\;\Conid{Π}\;(\Conid{El}\;\Varid{a}\;\Varid{pa})\;(\Conid{El}\;\Varid{b}\;\Varid{pb}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{tyOf}\;(\Varid{π[]}\;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\Varid{i})\;\mathrel{=}\;\Conid{U[]}\;\Varid{i}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

For example, the coherence proof for \ensuremath{\Varid{ηπ}} is given by
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{beginSub[}\;\Varid{ηπ}\;\mskip1.5mu]\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{elimSub}\;\Varid{σ}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{≡Sub[}\;\Varid{ηπ}\;\Varid{]⟨}\;\Varid{ηπ∙}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{⟩}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π₁}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{,}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π₂}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{tyOfπ₂}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\kern-1.5pt]^{\bullet}_{\text{t}}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{≡Sub[}\;\Varid{refl}\;\Varid{]⟨}\;\Varid{cong}\;{}\<[25]%
\>[25]{}(\Varid{λ}\;\Varid{z}\;\Varid{→}\;\Varid{...}\;\Varid{,}\;\Varid{...}\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{z}\;\kern-1.5pt]^{\bullet}_{\text{t}})\;{}\<[E]%
\\
\>[25]{}(\Conid{Ty∙-is-set}\;{}\<[38]%
\>[38]{}\anonymous \;\anonymous )\;\Varid{⟩}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{π₁}^{\bullet}\;(\Varid{elimSub}\;\Varid{σ})\;\Varid{,}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{elimTm}\;(\Varid{π₂}\;\Varid{σ})\;\Varid{∶[}\;\Varid{refl}\;\Varid{,}\;\Varid{elimTyOf}\;(\Varid{π₂}\;\Varid{σ})\;\Varid{refl}\;\kern-1.5pt]^{\bullet}_{\text{t}}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{∎}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
with the steps \ensuremath{\Varid{σ}^{\bullet}\;\Varid{≡Sub[}\;\Varid{p}\;\Varid{]⟨}\;\Varid{p}^{\bullet}\;\Varid{⟩}\;\Varid{τ}^{\bullet}} implemented using \ensuremath{\_\Varid{∙}_{\mathsf{P}}\_}.
These steps give us an equation over \ensuremath{\Varid{ηπ}\;\Varid{∙}\;(\Varid{refl}\;\Varid{∙}\;\Varid{refl})}, and \ensuremath{\Varid{beginSub[}\;\Varid{ηπ}\;\Varid{]\char95 }} gives us an easy way to correct this to an equation over \ensuremath{\Varid{ηπ}} as desired instead.
